---
title: "Introduction to cohortBuilder"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to cohortBuilder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options("tibble.print_min" = 5, "tibble.print_max" = 5)
library(magrittr)
library(cohortBuilder)
```

This document presents to you basic functionality offered by `cohortBuilder` package.
You'll learn here about Source and Cohort objects, how to configure them with filters 
and filtering steps.
Later on, we'll present most common Cohort methods that allow to manipulate the object and 
extract useful information about Cohort data and state.

## cohortBuilder vs. dplyr

If you're familiar with `dplyr` (or any other data manipulation package) you
may be wondering what `cohortBuilder` has been created for.

Our main goal for creating `cohortBuilder` was to provide a common syntax for operating (filtering)
on any data source you need. 
This follows the idea for having `dplyr` and its database counterpart `dbplyr` package.

In order to achieve the goal, we put an emphasis on possibility to write custom extensions 
in terms of data source type, or operating backend (underneath `cohortBuilder` uses `dplyr` 
to operate on data frames, but you may create an extension using e.g. `data.table`). 
See `vignette("custom-extensions")`.

The second goal was integration of `cohortBuilder` with `shiny`.
The GUI for `cohortBuilder` is provided by `shinyCohortBuilder` package.
With this extension you may easily open Cohort configuration panel locally,
or include it in you custom dashboard.

## Data: librarian

To present `cohortBuilder`'s functionality we'll be operating on `librarian` dataset.
`librarian` is a list of four tables, storing a sample of book library management database.

```{r}
cohortBuilder::librarian
```

To learn more check `?librarian`.

## Source object

Every time you work with `cohortBuilder` the crucial part is to properly define the data
source with `set_source` function.
Source is an R6 object storing metadata about data and its origin. 
The metadata allows `cohortBuilder` to distinct what methods to use when performing operations on it.

To define a new source you need to provide data (connection).

Let's create now a new source storing `librarian` data.
To do so, we pass one obligatory parameter `dtconn` to `set_source` method.

`dtconn` stores data connection responsible for informing `cohortBuilder` on what data 
are we gonna work (and what extension to use, if any).

If you want to operate on R-loaded list of tables, provide `tblist` class object.
`tblist` is just a named list of data frames having `tblist` class.

**Note.** In order to create 'tblist' object use `tblist`, e.g. `tblist(mtcars, iris)`.
**Note.** In order to convert list of data frames to 'tblist' just use `as.tblist`.

```{r}
str(as.tblist(librarian), max.level = 1)
```

Let's proceed with creating the source:

```{r}
librarian_source <- set_source(
  as.tblist(librarian)
)
class(librarian_source)
```

To learn more about `set_source`'s arguments check `?set_source`.

## Cohort object

When `Source` object is ready, the next step is to create a `Cohort` object.
`Cohort` is again an R6 object, providing methods for operating on data included in `Source`.

`Cohort` is responsible in particular for:

- storing definitions of filters (and filtering steps),
- running filtering and keeping result of it,
- computing and caching filter and data statistics,
- keeping and updating filtering configuration state.

In the standard workflow we build `Cohort` on top of `Source`.
We achieve it with `cohort` function:

```{r}
librarian_cohort <- librarian_source %>% 
  cohort()
class(librarian_cohort)
```

With the existing `Cohort` we may get underlying data with `get_data`:

```{r}
get_data(librarian_cohort)
```

We'll present more methods in the next sections.

## Configuring and running filters

The next step in `cohortBuilder` workflow is configuration of filters.
Filters are responsible for providing necessary logic for performing related data filtering. 

The extensive description of filters can be found at `vignette("custom-filters")`.

The current version of `cohortBuilder` provides five types of build-in filters:

- **discrete** - return values (in column) matching provided set,
- **discrete_text** - return values based on provided comma separated values,
- **range** - return values within the provided range,
- **date_range** - range version for Date type data,
- **multi_discrete** - extended version of discrete filter working with multiple conditions.

Let's define discrete filter that will subset `books` table listing books written by Dan Brown.  
To do so, we have to define the following parameters calling `filter` function:

- `type` - type of the filter (one of the above),
- `dataset` - name of the dataset to apply the filter to,
- `variable` - name of the variable in `dataset` to apply the filter to,
- `value` - vector of values to be applied in filter.

So in our case:

```{r}
author_filter <- filter(
  "discrete",
  dataset = "books",
  variable = "author",
  value = "Dan Brown"
)
```

In order to add the filter to existing Cohort we may use `add_filter` method:

```{r}
librarian_cohort <- librarian_cohort %>%
  add_filter(author_filter)
```

Alternatively we may use `%->%` operator that calls `add_filter` underneath:

```{r}
librarian_cohort <- librarian_cohort %->%
  author_filter
```

Or define the filter while creating Cohort:

```{r}
librarian_cohort <- librarian_source %>% 
  cohort(
    author_filter
  )
```

There are much more options for defining filters. 
To learn more check `vignette("cohort-configuration")`.

**Note.** Cohort is an R6 object, so you may skip reassignment above.

For example:

```{r, eval = FALSE}
librarian_cohort %>%
  add_filter(author_filter)
```

will also work.

**Note.** To verify if the filter was configured properly just run:

```{r}
sum_up(librarian_cohort)
```

The output highlights list of configured filters along with their parameters.
You can see here the id attached to filter and some extra parameters such as `keep_na`
or `active` which we describe in the next sections.

More to that we can realize the filter was defined in the step with ID equals to 1.
That's because `cohortBuilder` allows to perform [multi-stage filtering](#multi-stage-filtering).

Let's get back to filtering the `books`. 
Configuring filters only adds proper metadata in the Cohort object, which means 
data filtering is not performed automatically. 
This allows to set the proper configuration first, and run calculation only once.

If you want to run data filtering, just call `run`:

```{r}
run(librarian_cohort)
```

Let's check if the operation worked fine by checking the resulting data:

```{r}
get_data(librarian_cohort)
```

If you want to run data filtering automatically when the filter is defined you can
set `run_flow = TRUE`:

```{r}
librarian_cohort <- librarian_source %>% 
  cohort() %>% 
  add_filter(author_filter, run_flow = TRUE)
```

when using `add_filter` or:

```{r}
librarian_cohort <- librarian_source %>% 
  cohort(
    author_filter,
    run_flow = TRUE
  )
```

when configuring filter along with creating cohort.

Now when the data filtered, how can we get data state before filtering?
With `get_data` it's easy, just set `state = "pre"`:

```{r}
get_data(librarian_cohort, state = "pre")
```

## Multi-stage filtering

With `cohortBuilder` you can define filters in groups named 'steps' or 'filtering steps'.

Filtering steps allow you to sequentially perform groups of filtering operations.
In order to define step, just wrap set of filters in `step` function.

We will define three filters:

1. Taking all the books written by Dan Brown.
2. Filtering only the members (borrowers) with "standard" program.
3. Taking only the books with less than 5 copies.

We'll include filters 1. and 2. in the first step - filter 3. in the second one.

The below code does the job:

```{r}
librarian_cohort <- librarian_source %>% 
  cohort(
    step(
      filter(
        "discrete", id = "author", dataset = "books", 
        variable = "author", value = "Dan Brown"
      ),
      filter(
        "discrete", id = "program", dataset = "borrowers", 
        variable = "program", value = "premium", keep_na = FALSE
      )
    ),
    step(
      filter(
        "range", id = "copies", dataset = "books", 
        variable = "copies", range = c(-Inf, 5)
      )
    )
  )
```


Let's note a few parts that occurred above:

- For each filter we defined `id` parameter. 
This assigns provided id to each filter what makes accessing it later much easier.
- For 'program' filter we set `keep_na = FALSE` what results with excluding `NA` values 
(the parameter is available for each filter type).
- For filtering number of copies we've used `range` filter, for which sub-setting value 
is defined with `range` parameter.

Let's check the Cohort configuration:,

```{r}
sum_up(librarian_cohort)
```

We can see filters were correctly assigned to each step.

Having multiple steps defined, we can use `get_data` to extract resulting data after each step.
In order to precise the step we want to get data from, just pass its id as `step_id` parameter:

```{r}
run(librarian_cohort)
get_data(librarian_cohort, step_id = 1)
get_data(librarian_cohort, step_id = 2)
```

**Note.** When `step_id` is not provided, the method returns the last step data.

**Note.** You may precise if you want to extract data before or after filtering using `state` parameter.
Because the proceeding step uses result from the previous one, we have:

```{r}
identical(
  get_data(librarian_cohort, step_id = 1, state = "post"),
  get_data(librarian_cohort, step_id = 2, state = "pre")
)
```

## Exploring the Cohort object methods

### Learning more about the source data

Having Cohort object created, you may want to use its methods for exploring underlying data.

With methods such as:

- `stat`,
- `plot_data`,
- `attrition`

you can:

- get filter related data statistics,
- display filter related data on plot,
- display data changes across filtering steps.

```{r}
stat(librarian_cohort, step_id = 1, filter_id = "program")
stat(librarian_cohort, step_id = 2, filter_id = "copies")
```

```{r}
plot_data(librarian_cohort, step_id = 1, filter_id = "program")
```
```{r}
plot_data(librarian_cohort, step_id = 2, filter_id = "copies")
```

```{r}
attrition(librarian_cohort, dataset = "books")
```
```{r}
attrition(librarian_cohort, dataset = "borrowers")
```


### Sharing code and reproducibility

The `cohortBuilder` package offers some methods to make sharing the workflow easier.

With `code`, you may get the reproducible code written using methods operating on 
specific source (i.e. `dplyr` for `tblist` and `dbplyr` for `db` source):

```{r}
code(librarian_cohort)
```

We can see above, the resulting code uses `source` object, which creation can be
defined separately while creating it:

```{r}
librarian_source <- set_source(
  as.tblist(librarian),
  source_code = quote({
    source <- list(attributes = list(datasets = librarian))
  })
)

librarian_cohort <- librarian_source %>% 
  cohort(
    step(
      filter(
        "discrete", id = "author", dataset = "books", 
        variable = "author", value = "Dan Brown"
      ),
      filter(
        "discrete", id = "program", dataset = "borrowers", 
        variable = "program", value = "premium", keep_na = FALSE
      )
    ),
    step(
      filter(
        "range", id = "copies", dataset = "books", 
        variable = "copies", range = c(-Inf, 5)
      )
    ),
    run_flow = TRUE
  )

code(librarian_cohort)
```

The second option allows to restore cohort configuration using its state.
The cohort state is a list (or json) storing information about all the steps and filters configuration.

You may get the state with `get_state` method:

```{r}
state <- get_state(librarian_cohort, json = TRUE)
state
```

Then, having an empty cohort, use `restore` to apply the configuration:

```{r}
librarian_cohort <- librarian_source %>%
  cohort()

restore(librarian_cohort, state = state)

sum_up(librarian_cohort)
```
