---
title: "Writing custom extensions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing custom extensions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options("tibble.print_min" = 5, "tibble.print_max" = 5)
library(magrittr)
library(cohortBuilder)
```

`cohortBuilder` package is adapted to work with various data sources and custom backends.
Currently there exists one official extension `cohortBuilder.db` package that allows you to use `cohortBuilder` with database connections.

The goal of this document is to explain how to create custom extensions to `cohortBuilder`.

In general to create the custom layer you need to:
 
1. Define set of S3 methods operating on your custom source.
2. Define selected filters (see `vignette("custom-filters")`).

It's recommended to include all of the methods in your custom R package.

Before you start creating a new layer, you need to choose what data (connection) should your layer operate on.

For example, cohortBuilder uses `tblist` class object to operate on list of data frames , or `db` class for operating on database connections.

To start with create a function that will take required parameters to define data connection, such as `tblist` or `dbtables` in case of `cohortBuilder.db`. The function should return an object of selected class which is used to define required extension methods.

Below we describe all the required and optional methods you need to define within the created package.

1. **`set_source` - method used for defining a new source**

Required parameters:

- `dtconn`

Details:

- Define the method calling `Source$new` inside.
- Declare necessary parameters used for defining data source (and pass them to `Source$new`). The arguments are then available at source$attributes object.
- If valid, you may declare `primary_keys` and `binding_keys` parameters (see `vignette("binding-keys")`).
- It's also worth to declare `source_code` parameter that allows users to define code for creating source (visible in reproducible code) and `description` storing list of useful source objects descriptions.

Example:

- `cohortBuilder` - `tblist` object (same for `cohortBuilder.db` - `db` object)

```{r, eval = FALSE}
set_source.tblist <- function(dtconn, primary_keys = NULL, binding_keys = NULL, 
                           source_code = NULL, description = NULL, ...) {
  Source$new(
    dtconn, primary_keys = primary_keys, binding_keys = binding_keys,
    source_code = source_code, description = description,
    ...
  )
}
```

2. **`.init_step` - structure data passed between filtering steps**

Required parameters:

- `source` - Source object

Details:

- Within the function body define how data should be extracted from source, and structured.
- A structure of the output is then used as an input for filter's method (`data_object` argument).

Examples:

- `cohortBuilder` - 'tblist' class. Operating on list of tables in each step.

```{r, eval = FALSE}
.init_step.tblist <- function(source, ...) {
  source$dtconn
}
```

- `cohortBuilder.db` - 'db' class.

`cohortBuilder.db` operates on `db` class object which is list of `connection`, `tables` and `schema` fields. 

```{r, eval = FALSE}
.init_step.db <- function(source) {
  purrr::map(
    stats::setNames(source$dtconn$tables, source$dtconn$tables),
    function(table) {
      tbl_conn <- dplyr::tbl(
        source$dtconn$connection,
        dbplyr::in_schema(source$dtconn$schema, table)
      )
      attr(tbl_conn, "tbl_name") <- table
      tbl_conn
    }
  )
}
```

3. **`.pre_filtering` (optional) - modify data object before filtering**

Required parameters:

- `source`, 
- `data_object` - an object following the structure of `.init_step`, 
- `step_id` - id of the filtering step

Details:

- An optional method used to prepare data for filtering (cleaning up attributes, creating middle step objects etc.).

Examples:

- `cohortBuilder` - tblist class. Cleaning up `filtered` attribute for new step data.

```{r, eval = FALSE}
.pre_filtering.tblist <- function(source, data_object, step_id) {
  for (dataset in names(data_object)) {
    attr(data_object[[dataset]], "filtered") <- FALSE
  }
  return(data_object)
}
```

- `cohortBuilder.db` - creating temp tables for the current step in database and cleaning up `filtered` attributes.

```{r, eval = FALSE}
.pre_filtering.db <- function(source, data_object, step_id) {
  purrr::map(
    stats::setNames(source$dtconn$tables, source$dtconn$tables),
    function(table) {
      table_name <- tmp_table_name(table, step_id)
      DBI::dbRemoveTable(source$dtconn$conn, table_name, temporary = TRUE, fail_if_missing = FALSE)
      attr(data_object[[table]], "filtered") <- FALSE
      return(data_object[[table]])
    }
  )
}
```

4. **`.post_filtering` (optional) - data object modification after filtering (before running binding).**

Required parameters:

- source, 
- data_object - an object following the structure of `.init_step`, 
- step_id - id of the filtering step

5. **`.post_binding` (optional) - data object modification after running binding.**

Required parameters:

- source, 
- data_object - an object following the structure of `.init_step`, 
- step_id - id of the filtering step

6. **`.collect_data` - define how to collect data object into R.**

Required parameters:

- source, 
- data_object - an object following the structure of `.init_step`

Details:

- `cohortBuilder`'s equivalent of `collect` method known for sourcing the object into R memory when working with remote environment (e.g. database).
- When operating in R memory it's enough to return `data_object`.

Examples:

- `cohortBuilder` - operating in R memory, so return `data_object`.

```{r, eval = FALSE}
.collect_data.tblist <- function(source, data_object) {
  data_object
}
```

- `cohortBuilder.db` - collect tables from database and return as a named list.

```{r, eval = FALSE}
.collect_data.db <- function(source, data_object) {
  purrr::map(
    stats::setNames(source$dtconn$tables, source$dtconn$tables),
    ~dplyr::collect(data_object[[.x]])
  )
}
```

7. **`.get_stats` - collect data object stats**

Required parameters:

- `source`, 
- `data_object`

Details:

- There are no obligatory statistics to be returned.
- The calculated statistic is cached within Cohort object after each step filtering.
- Returned stats plays mostly read-only role. The only situation the stats are used is within `.get_attrition_count` and `shinyCohortBuilder` integration.

Examples:

- `cohortBuilder` - operating in R memory, so return `data_object`.

```{r, eval = FALSE}
.get_stats.tblist <- function(source, data_object) {
  dataset_names <- names(source$dtconn)
  dataset_names %>%
    purrr::map(
      ~ list(n_rows = nrow(data_object[[.x]]))
    ) %>%
    stats::setNames(dataset_names)
}
```

- `cohortBuilder.db` - collect tables from database and return as a named list.

```{r, eval = FALSE}
.get_stats.db <- function(source, data_object) {
  dataset_names <- source$dtconn$tables
  dataset_names %>%
    purrr::map(
      ~ list(
        n_rows = data_object[[.x]] %>%
          dplyr::summarise(n = n()) %>%
          dplyr::collect() %>%
          dplyr::pull(n) %>%
          as.integer()
      )
    ) %>%
    stats::setNames(dataset_names)
}
```

8. **`.run_binding` - method defining how [binding](binding-keys.html) should be handled**

Required parameters:

- `source`, 
- `binding_key` - binding key definition, 
- `data_object_pre` - data object state before filtering in the current step, 
- `data_object_post` - data object state after filtering in the current step (including effect of previous bindings)

Details:

- Binding is run after applying all the filters in the current filtering step (and applying `.post_filtering` if defined).
- When filtering is finished, binding loop starts iterating over all the defined bindings. `.run_binding` takes care of handling a single iteration.
- The returned bound data, should preserve the structure of `.init_step` method output.
- You should preserve the assumed binding rules described in [bindings-keys](binding-keys.html) (i.e. handling `post = TRUE/FALSE`, `activate = TRUE/FALSE` and `filtered` attribute) but this is not obligatory.

Examples:

- `cohortBuilder`

```{r, eval = FALSE}
.run_binding.tblist <- function(source, binding_key, data_object_pre, data_object_post, ...) {
  binding_dataset <- binding_key$update$dataset
  dependent_datasets <- names(binding_key$data_keys)
  active_datasets <- data_object_post %>%
    purrr::keep(~ attr(., "filtered")) %>%
    names()

  if (!any(dependent_datasets %in% active_datasets)) {
    return(data_object_post)
  }

  key_values <- NULL
  common_key_names <- paste0("key_", seq_along(binding_key$data_keys[[1]]$key))
  for (dependent_dataset in dependent_datasets) {
    key_names <- binding_key$data_keys[[dependent_dataset]]$key
    tmp_key_values <- dplyr::distinct(data_object_post[[dependent_dataset]][, key_names, drop = FALSE]) %>%
      stats::setNames(common_key_names)
    if (is.null(key_values)) {
      key_values <- tmp_key_values
    } else {
      key_values <- dplyr::inner_join(key_values, tmp_key_values, by = common_key_names)
    }
  }

  data_object_post[[binding_dataset]] <- dplyr::inner_join(
    switch(
      as.character(binding_key$post),
      "FALSE" = data_object_pre[[binding_dataset]],
      "TRUE" = data_object_post[[binding_dataset]]
    ),
    key_values,
    by = stats::setNames(common_key_names, binding_key$update$key)
  )
  if (binding_key$activate) {
    attr(data_object_post[[binding_dataset]], "filtered") <- TRUE
  }

  return(data_object_post)
}
```

- `cohortBuilder.db` - slight modification of the above function

9. **`.get_attrition_count` - define how to get metric used for attrition data plot**

Required parameters:

- `source`, 
- `data_stats` - statistics related to each step data - list of `.get_stats` results for each step (and original data, assigned to `step_id = 0`), 

Details:

- The output should return vector of length `n+1` where `n` is number of steps. The first element of the vector should describe statistic for the base, unfiltered data.
- You may define additional parameters which can be passed to `attrition` method of Cohort object (e.g. `dataset` in the below example).

Examples:

- `cohortBuilder`

```{r, eval = FALSE}
.get_attrition_count.tblist <- function(source, data_stats, dataset, ...) {
  data_stats %>%
    purrr::map_int(~.[[dataset]][["n_rows"]])
}
```

- `cohortBuilder.db` - same as above

10. **`.get_attrition_label` - define label displayed in attrition plot for the specified step**

Required parameters:

- `source`, 
- `step_id` - id of the step (`"0"` for original data case),
- `step_filters` - list storing filters configuration for the selected step (`NULL` for original data case),

Details:

- Please remember to define label for the initial data state (`step_id = "0"` case).
- You may define additional parameters which can be passed to `attrition` method of Cohort object (e.g. `dataset` in the below example).

Examples:

- `cohortBuilder`

```{r, eval = FALSE}
get_attrition_label.tblist <- function(source, step_id, step_filters, dataset, ...) {
  pkey <- source$primary_keys
  binding_keys <- source$binding_keys
  if (step_id == "0") {
    if (is.null(pkey)) {
      return(dataset)
    } else {
      dataset_pkey <- .get_item(pkey, "dataset", dataset)[1][[1]]$key
      if (is.null(dataset_pkey)) return(dataset)
      return(glue::glue("{dataset}\n primary key: {paste(dataset_pkey, collapse = ', ')}"))
    }
  }
  filters_section <- step_filters %>%
    purrr::keep(~.$dataset == dataset) %>%
    purrr::map(~get_attrition_filter_label(.$name, .$value_name, .$value)) %>%
    paste(collapse = "\n")
  bind_keys_section <- ""
  if (!is.null(binding_keys)) {
    dependent_datasets <- .get_item(
      binding_keys, attribute = "update", value = dataset,
      operator = function(value, target) {
        value == target$dataset
      }
    ) %>%
      purrr::map(~names(.[["data_keys"]])) %>%
      unlist() %>%
      unique()
    if (length(dependent_datasets) > 0) {
      bind_keys_section <- glue::glue(
        "\nData linked with external datasets: {paste(dependent_datasets, collapse = ', ')}",
        .trim = FALSE
      )
    }
  }
  gsub(
    "\n$",
    "",
    glue::glue("Step: {step_id}\n{filters_section}{bind_keys_section}")
  )
}
```

- `cohortBuilder.db` - same as above
