---
title: "Source intelligence and AI tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Source intelligence and AI tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options("tibble.print_min" = 5L, "tibble.print_max" = 5L)
library(cohortBuilder)
```

Beyond manually configuring filters, `cohortBuilder` can inspect a source and
describe or build filters for you. These features are also the foundation for
integrating a cohort with a Large Language Model (LLM), so an assistant can
explore the data and apply filters on the user's behalf.

This article covers four building blocks:

- `describe()` - attach human-readable descriptions to datasets and variables,
- `autofilter()` - auto-generate filters from the data structure,
- `shape()` - return a structured summary of datasets and filters,
- AI tools (`R/ai_tools.R`) - expose the cohort to an `ellmer` chat.

## Describing a source

`describe()` builds a small description object (its text plus any extra fields)
that you attach to a source via the `description` argument of `set_source()`.

The `description` is a nested list keyed by dataset name. Within each dataset,
the special key `dataset_` describes the dataset itself, and any other key
describes a variable of that dataset:

```{r}
iris_source <- set_source(
  tblist(iris = iris),
  description = list(
    iris = list(
      dataset_ = describe("Edgar Anderson's measurements of iris flowers."),
      Species = describe("Iris species.", domain = c("setosa", "versicolor", "virginica"))
    )
  )
)
```

Extra named arguments to `describe()` (such as `domain` above) are stored
alongside the text and can be picked up by other features - for example
`autofilter()` uses a supplied `domain` instead of scanning the data.

## Generating filters automatically

`autofilter()` analyses each column of the source and creates a filter suited
to its type (using filter rules such as `rule_character`, `rule_factor`,
`rule_numeric`, `rule_Date`, `rule_POSIXct`). The mapping is roughly:

| Column type | Filter type |
|-------------|-------------|
| character / factor | `discrete` (or `discrete_text` when all values are unique) |
| numeric / integer  | `range` |
| Date               | `date_range` |
| POSIXct            | `datetime_range` |

The `attach_as` argument controls where the generated filters go.

With `attach_as = "step"` (the default) the filters are added as a filtering
step, so the cohort is immediately filterable:

```{r}
iris_cohort <- set_source(tblist(iris = iris)) |>
  autofilter(attach_as = "step") |>
  cohort()

sum_up(iris_cohort)
```

With `attach_as = "meta"` the filters are stored in `source$available_filters`
rather than applied. This is the "menu" of filters a GUI or an LLM can choose
from, without forcing them onto the data:

```{r}
meta_source <- iris_source |>
  autofilter(attach_as = "meta")

length(meta_source$available_filters)
```

When a `domain` was provided via `describe()`, the generated filter inherits it
instead of scanning the data:

```{r}
species_filter <- purrr::detect(
  meta_source$available_filters, ~ .x@id == "iris-Species"
)
species_filter@domain
```

## Inspecting a source with `shape()`

`shape(source)` returns a structured `list(datasets, filters)` describing the
source - ideal for programmatic inspection or passing to an LLM:

- `datasets` maps each dataset name to its description text (or `NA`).
- `filters` is keyed by filter id; each entry is a list with `dataset`, `type`,
  `description`, `variables`, and `domain`.

```{r}
result <- shape(meta_source)

# Dataset descriptions
result$datasets

# One filter entry
str(result$filters$`iris-Species`)
```

When a filter's own `@domain` is unset, `shape()` falls back to the domain
stored in the source's metadata statistics, so the `domain` field is populated
whenever possible.

**Note.** Called with a `field` (and optional `subfield`),
`shape(source, field, subfield)` instead performs a description-text lookup -
this is the form used internally by `Cohort$show_help()`.

## Connecting a cohort to an LLM

The functions in `R/ai_tools.R` wrap cohort operations as tools an
[`ellmer`](https://ellmer.tidyverse.org/) chat can call. Each tool is a
`cb_tool` object (a function plus a name, description, and argument schema).

The built-in tool factories each take a cohort and return a `cb_tool`:

| Tool factory | Purpose |
|--------------|---------|
| `cb_tool_filters_meta()` | Return available-filter metadata (via `shape()`) as JSON |
| `cb_tool_describe_state()` | Describe current steps, filters, and pending state |
| `cb_tool_get_data_summary()` | Report row counts per dataset and step |
| `cb_tool_get_code()` | Return reproducible filtering code |
| `cb_tool_add_filters()` | Add filters (no values) to a new or existing step |
| `cb_tool_set_filter_values()` | Set values on existing filters |
| `cb_tool_apply_filters()` | Add filters and set their values in one call |
| `cb_tool_toggle_filters()` | Activate / deactivate filters |
| `cb_tool_clear_filters()` | Reset filters to their defaults |
| `cb_tool_remove_filters()` | Remove filters from a step |
| `cb_tool_remove_step()` | Remove the last step |
| `cb_tool_run()` | Run the pipeline (when auto-run is disabled) |

A `cb_tool` prints its name, description, and arguments:

```{r}
coh <- cohort(meta_source)
tool <- cb_tool_filters_meta(coh)
print(tool)
```

For LLM-driven filtering to work, the source must expose a menu of filters via
`autofilter(attach_as = "meta")` so the assistant knows what it can apply.

To register tools with an `ellmer` chat, use `cb_register_tool()` for a single
tool or `cb_register_tools()` to register all of them at once:

```{r, eval = FALSE}
library(ellmer)

source <- set_source(tblist(iris = iris)) |>
  autofilter(attach_as = "meta")
coh <- cohort(source)

chat <- chat_openai()
chat |> cb_register_tools(coh)

chat$chat("Filter the data to setosa flowers with sepal length over 5")
```

By default the cohort runs automatically after each tool modifies it. Set
`options(cb_tool_run_cohort = FALSE)` to require an explicit `cb_run` call
instead.

**Note.** The AI tools require the suggested `ellmer` package.
